Extending Height Above Nearest Drainage to Model Multiple Fluvial Sources in Flood Inundation Mapping Applications for the U.S. National Water Model

### 2022-08-21 manuscripts/owp_fim4_2022/sections/introduction.tex
* Draft containing all revisions after first round of reviews * This version was submitted to co-authors for comments

With respect to property damages, river and flash flooding have contributed to 60.7, 1.6, and 3.7 billion non-inflation adjusted US dollars in the annual periods of 2017 to 2019, respectively \cite{national_weather_service_2020,national_weather_service_2019,national_weather_service_2018} with the large spike in 2017 attributed to the Hurricane Harvey event along the Gulf coast. 
Trends related to flood damages and fatalities have been steadily increasing over recent decades. \cite{mallakpour2015changing,downton2005reanalysis,kunkel1999temporal,pielke2000precipitation,corringham2019effect}. 
Some are expecting that the hydrologic cycle will intensify which will lead to more extreme precipitation in some areas along with a greater risk of flooding \cite{tabari2020climate,milly2002increasing,wing2018estimates}. 
Increasing trends in frequency and risk are not uniform across spatial regions with work by \citeA{slater2016recent} indicating that trends are increasing across the US Midwest/Great Lakes region while decreasing in coastal Southeast, Southwest and California.

@@

\caption{Forecast points with and without FIM in United States' Advanced Hydrologic Prediction System.
Also shown are the National Water Model stream networks at the full resolution (FR) and Mainstems (MS) resolution.}

@@

Additional work is required to fill-in the gaps that the AHPS leaves in terms of spatial and temporal coverage.
In response to growing stakeholder demand for enhanced and integrated water resource forecasts, the Office of Water Prediction (OWP) at the National Water Center (NWC) along with its partners has developed and implemented operationally the National Water Model (NWM) which is a version of the Weather Research and Forecast Hydrologic Model (WRF-Hydro) \cite{gochis2018wrf,cosgrove2019evolution}. 
The NWM forecasts river discharges at more than 2.7 million forecast points at a variety of time horizons including some medium (10 day) and long (30 day) range forecast horizons.

@@

As a complement to the operational NWM, OWP has also developed a configuration of the NWM that extends RFC forecasts downstream by assimilating and routing forecast streamflows to the next downstream AHPS forecast point.
This configuration of the NWM is used to enhance forecasting skill by leveraging best available regional-scale data.
The river network upon which this special configuration operates is herein referred to as the Main-Stem (MS) modeling stream network.
Figure \ref{ssec:national_water_model} shows the Full-Resolution (FR) modeling stream network as well as the MS network.

@@

The National Hydrography Dataset Plus (NHDPlus) V2.1 is the basis for the hydrofabric in the NWM due to its comprehensive use with the hydrologic communities' stakeholders \cite{mckay2012nhdplus}. 
The term hydrofabric is used within the NWM jargon to describe the subset of hydrography composed of the geospatial datasets required for hydrologic modeling including but not limited to stream networks, catchments, channel properties, and elevation data. 
The Muskingam-Cunge routing method is used within the NWM to reduce computational requirements of a continental scale model \cite{bedient2008hydrology,ponce1994variable,gochis2018wrf}.
Muskingam-Cunge routing scheme has been demonstrated by \citeA{cunge1969subject} to be equivalent to the convective-diffusive wave method without consideration to wave dampening.
As a result of high computational costs and large spatial domains, the need for high-resolution FIM at 10 m or better requires additional post-processing from the principal output of the NWM which is forecast river discharges at the reach scale. The Height Above Nearest Drainage (HAND) terrain model is one such technique that can be used, along with synthetic rating curves (SRC), to convert riverine discharges to stages then finally to inundation extents.

@@

Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEMs such as the 3D Elevation Program (3DEP) seamless at the 1/3 arc-second (10 m) scale, HAND has been implemented into software for large-scale, continental computation. 
As part of the OWP's Innovators Program and NWC's Summer Institute, the National Flood Interoperability Experiment (NFIE) generated FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a high-performance computer (HPC) \cite{maidment2017conceptual,liu2016cybergis}. 
NFIE used open-source dependencies including the Terrain Analysis Using Digital Elevation Models (TauDEM) \cite{tarboton2005terrain} and the Geospatial Data Abstraction Library (GDAL) \cite{warmerdam2008geospatial} to compute HAND for the Continental United States (CONUS) at 331 Hydrologic Unit Code (HUC) 6 processing units in 1.34 CPU years.
By allocating 31 nodes at 20 cores per for a total of 620 available cores to the overall operation, it enabled the production to finish up in 36 hours consuming 3.2TB of peak memory and 5TB of total disk space.
Originally, NFIE utilized the NHD Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (HR) \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC as an unofficial demonstrations to produce forecast FIM from the NWM for additional guidance in hydro-blind regions.
Further work by \citeA{djokic2019arc}, implemented a series of improvements to HAND including equidistant reaches, updates to use with NHDPlusHR hydrography, and AGREE-DEM reconditioning \cite{hellweger1997agree} into an ESRI Arc-Hydro workflow with use in ArcGIS. 
More notably the software added the ability to derive drainage potentials on both the NWM FR and MS stream networks which leverages the lower drainage density and Horton-Strahler stream order of the MS network.

Related to these efforts, the USGS has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses synthetic rating curves with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{OWP FIM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Many of those in the academic community assessing HAND's efficacy for producing FIM have noted opportunities for improvement. 
\citeA{godbout2019error} found how reach length and slope are important parameters for maximizing mapping skill with the moderate values performing best. 
The co-linearity of reach length and slope led \citeA{godbout2019error} to propose that reaches of extreme lengths performed worse because of the extreme slope values, a parameter directly represented in Manning's equation. 
Issues with the reach-average approaches have been documented in \citeA{tuozzolo2019impact} where large within reach variance of the roughness Manning's n coefficient have been observed.
Furthermore, \citeA{garousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3 m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).
The use of higher resolution DEMs in that study was motivated by previous work with Lidar DEMs and least-cost thalweg derivations \cite{zheng2018geoflood}.
Further work by \citeA{johnson2019integrated} noted the general under-prediction of HAND and suggested tuning the Manning's n parameter to better align SRC's with observations. 
Additionally, the sensitivity to low topographic relief and channel slope have been observed \cite{johnson2019integrated,godbout2019error}. 
Most notably, HAND has been shown to demonstrate sensitivity to drainage network density known colloquially as the catchment boundary problem \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is noted at higher order streams with large flows and low Froude numbers and is manifested by a constriction in the inundation extents at the junction location.
These findings suggest improvements to HAND are required that utilize advanced computational algorithms and software to compute a FIM hydrofabric required for producing continental-scale FIM.

With all of the latest developments in the realm of continental FIM (CFIM) noted in the previous paragraph, a fast, open-source framework for accelerating the research to operations pipeline is required.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
This framework is open-source and modular enabling and accelerating the research to operations development pipeline.
Automated evaluation tools and processed test case data enable the rapid testing and evaluation of new techniques in consistent contexts.
In addition to developed tooling, we introduce research demonstrating how FIM performance skill with HAND can be improved by reducing Horton-Strahler stream orders \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} of the stream networks.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to stream threshold which serves as a proxy for stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths, can enhance FIM skill.
This capability will be referred to as Generalized Mainstems (GMS).
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM with applications.
%

>>>>

With respect to property damages, river and flash flooding have contributed to 60.7, 1.6, and 3.7 billion non-inflation adjusted US dollars in the annual periods of 2017 to 2019, respectively \cite{national_weather_service_2020,national_weather_service_2019,national_weather_service_2018}, with the large spike in 2017 attributed to the Hurricane Harvey event along the Gulf Coast. 
Trends related to flood damages and fatalities have been steadily increasing over recent decades \cite{mallakpour2015changing,downton2005reanalysis,kunkel1999temporal,pielke2000precipitation,corringham2019effect}. 
Some are expecting that the hydrologic cycle will intensify due to climate change which will lead to more extreme precipitation in some areas along with a greater risk of flooding \cite{tabari2020climate,milly2002increasing,wing2018estimates}. 
Increasing trends in frequency and risk are not uniform across spatial regions with work by \citeA{slater2016recent} indicating that trends are increasing across the US Midwest and Great Lakes regions while decreasing in the coastal Southeast, Southwest, and California.

@@

\caption{Forecast points with and without Flood Inundation Maps (FIM) in United States' Advanced Hydrologic Prediction System (AHPS).
Note that only a small fraction of the AHPS forecast points have existing FIM.
Also shown are the National Water Model (NWM) stream networks at the Full Resolution (FR) and Mainstems (MS) resolution.
The FR network constitutes the entire NWM stream network while the MS resolution network is the FR network at or downstream of the AHPS forecast points shown.
}

@@

Additional work is required to address the gaps that the AHPS leaves in terms of spatial resolution, spatial coverage, and temporal forecast horizons.
In response to growing stakeholder demand for enhanced and integrated water resource forecasts, the Office of Water Prediction (OWP) at the National Water Center (NWC) along with its partners at the National Center for Atmospheric Research (NCAR) have developed and implemented operationally the NWM which is a configuration of the Weather Research and Forecasting Hydrologic Model (WRF-Hydro) \cite{gochis2018wrf,cosgrove2019evolution}. 
The NWM forecasts river discharges at more than 2.7 million forecast points at a variety of time horizons including lookback-range (3-28 hrs), short-range (18 hr), medium-range (10 day) and long-range (30 day) forecast horizons.

@@

As a complement to the operational NWM, RFC forecasts from AHPS forecast points are assimilated in the NWM and routed downstream to the next downstream AHPS forecast point where the process iterates again.
This assimilation into NWM is used to enhance forecasting skill by leveraging best available regional-scale forecasts.
The river network upon which this special assimilation technique operates on is herein referred to as the Mainstem (MS) stream network.
Figure \ref{fig:forecast_points} shows the NWM V2.1 FR stream network as well as the NWM V2.1 MS network.

@@

The National Hydrography Dataset Plus (NHDPlus) V2.1 is the basis for the ``hydrofabric'' in the NWM due to its comprehensive use with the hydrologic communities' stakeholders \cite{mckay2012nhdplus,nhdplus2022vectors}. 
The term ``hydrofabric'' is used within the NWM jargon to describe the subset of hydrography composed of the geospatial datasets required for hydrologic modeling including but not limited to stream networks, catchments, channel properties, and elevation data. 
The NWM provides stream forecasts at these hydrofabric segments using the Muskingum-Cunge method to reduce computational requirements of a continental scale model but fails to consider backwater dynamics \cite{bedient2008hydrology,ponce1994variable,gochis2018wrf}.
The need for high-resolution FIM at 10 m or better requires additional post-processing from the principal output of the NWM which is forecast river discharges at the reach scale.
The use of a 2-dimensional (2D) hydrodynamic model across a continental-scale and high spatial resolutions is very cost prohibitive especially in an operational setting.
The Height Above Nearest Drainage (HAND) terrain model is one such technique that can be used, along with synthetic rating curves (SRC), to convert 1-dimensional (1D) riverine discharges to stages, and finally to inundation extents and depths.

@@

HAND operates on many underlying assumptions since it can only be used as an inundation proxy or no physics model and thus, not a true hydrodynamic inundation model \cite{nobre2016hand,liu2016cybergis,liu2020height}.
HAND, to our knowledge, has only been applied to natural, inland, and riverine inundation applications thus it is also missing pluvial, coastal, ground water, and dam break components among other possible sources of flooding.
Additionally, in order to flood an area, HAND assumes all areas eligible for inundation must drain to some nearest stream line which is used for catchment allocation and relative elevation calculation \cite{nobre2016hand,nobre2011height,liu2016cybergis,liu2020height,maidment2017conceptual,garousi2019terrain,zheng2018river,zheng2018geoflood,johnson2019integrated,renno2008hand}.
Stream thalweg networks must also collectively drain to a singular outlet point for a given processing region \cite{nobre2016hand,zheng2018geoflood,renno2008hand}.
Since elevations don't naturally do this, they must undergo a long series of hydro-conditioning processes to enforce monotonically decreasing elevations across an entire processing unit along with hydrologically correct directions of flow \cite{nobre2016hand,nobre2011height,liu2016cybergis,liu2020height,donchyts2016global,renno2008hand}.
The level of DEM manipulation required to enforce this assumption can be substantial depending on the region and can be a significant source of error.
The drainage enforcing assumption also interacts with an inability to properly account for fluvial inundation in regions of DEM depressions that lack natural drainage to riverine areas \cite{nobre2016hand,renno2008hand}.

When used for FIM applications, HAND assumes only fluvial inundation sourced from its nearest drainage line is accounted for \cite{nobre2016hand,mcgehee2016modified}.
Catchments are independent of one another for FIM purposes meaning a reaches' stage value is only used to threshold the HAND values within its respective catchment \cite{liu2016cybergis,zheng2018river,zheng2018geoflood}.
This assumption plays to the ``Nearest Drainage'' term in HAND and creates a significant limitation within HAND for FIM applications \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
At the junction of high stream order and high flow rivers with lower flow tributaries, there can be a lack of inundation extents exhibited which is known colloquially in the forecasting community as the ``catchment boundary problem''.
The academic community has somewhat referenced this issue before but it has been characterized more as a problem with the stream delineation process that comes from thresholding the drainage accumulation maps \cite{nobre2016hand,li2020evaluation}.
Later in this study, we will re-introduce this problem and demonstrate how we initialize with a stream network (that of the NWM's) and thus avoid having to threshold accumulations to some arbitrary value to define stream networks.
We illustrate how computing HAND independently for stream lines of unit stream order can significantly enhance FIM performance by accounting for multiple sources of fluvial inundation that may exist in certain regions and flow scenarios.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{HAND Implementations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEMs such as the 3D Elevation Program (3DEP) seamless at the 1/3 arc-second (approximately 10 m depending on latitude) scale, HAND has been implemented into software for large-scale, continental computation. 
As part of the OWP's Innovators Program and NWC's Summer Institute, the National Flood Interoperability Experiment (NFIE) generated FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a HPC \cite{maidment2017conceptual,liu2016cybergis}. 
NFIE used open-source dependencies including the Terrain Analysis Using Digital Elevation Models (TauDEM) \cite{tarboton2005terrain} and the Geospatial Data Abstraction Library (GDAL) \cite{warmerdam2008geospatial} to compute HAND for the Continental United States (CONUS) at 331 Hydrologic Unit Code (HUC) 6 processing units in 1.34 CPU years.
By allocating 31 nodes at 20 cores per node for a total of 620 available cores to the overall operation, it enabled the production to finish up in 36 hours consuming 3.2TB of peak memory and 5TB of total disk space.
Originally, NFIE utilized the NHD Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (NHDPlusHR) which better agrees with the 10 m DEM from the NHDPlusHR program \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC as an unofficial demonstration to produce forecast FIM from the NWM for additional guidance in hydro-blind regions.
Further work by \citeA{djokic2019arc}, implemented a series of improvements to HAND including equidistant reaches, updates to use with NHDPlusHR hydrography, and AGREE DEM reconditioning \cite{hellweger1997agree} into an ESRI Arc-Hydro workflow with use in ArcGIS. 
More notably the software added the ability to derive HAND on both the NWM FR and MS stream networks to consider multiple sources of fluvial inundation along high impact rivers of primary forecasting concern.

Related to these efforts, the United States Geological Survey (USGS) has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses SRCs with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Office of Water Prediction Flood Inundation Mapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In order to mitigate the ever increasing threat of flooding to life and property, an operational capability is required to extend NWM streamflow forecasts to river stages, inundation extents, and inundation depths.
OWP FIM is introduced here as a continental scale capability that generates these products at high spatial and temporal resolutions.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
Within the operational framework of OWP FIM, we introduce research demonstrating how FIM performance skill with HAND can be improved by discretizing stream networks into units of an effective unit Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} for HAND computation contexts.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to the stream accumulation threshold which is closely related to stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths (LP), can enhance FIM skill by accounting for multiple possible sources of fluvial inundation.
This capability is introduced progressively as MS (whose network represents about 4\% of FR network) and to a higher degree Generalized Mainstems (GMS) (covers entire FR network) which will be explained later on.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM with applications.
%
